---
title: "Tissue Specific Gene Analysis"
author: "Adam Gruenbaum"
date: "September 25, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

First we read the results from bcbio into R. The data.frame package provides a variety of tools for working with data frames in R. The `fread()` function is a significantly more efficient and flexible tool for reading in tables than base R, so we'll use it. The annotated_combined.count will be in samples/final/ and then within a folder with the date of the run. Put the name of this folder in folder_name.

```{r read data}
# import data.table library
library(data.table)
# name of folder containing combined count table 
# WILL NEED TO BE UPDATED WITH DATE OF BCBIO RUN
folder_name = "2019-09-12_bcbio_samples"
# make full path of annotated_combined count
count_path = paste0("bcbio_samples/final/", folder_name, "/annotated_combined.counts")
# read table
data <- data.frame(fread(count_path))
rownames(data) = data$id
data$symbol = data$id = NULL
```

We'd next like to filter our data a little. We'll remove genes that don't pass a given threshold of counts per million in at least n samples. We'll use the edgeR package for this. edgeR is one of a few different packages for analysing RNAseq data, including DESeq2 and limma. All have their merits and the exact package of choice is as much personal preference as anything else. We use limma for our actual analysis, but edgeR’s tools for normalizing the data and filtering it are useful. 

```{r fitler genes}
# genes must have at least 50 cpm
cpm_threshold = 50
# in at least 2 samples
cpm_needed = 2

#isolate just 'Ctrl' or 'L' from colnames
tissues = factor(sub('_.*_.*', '', colnames(data)))

# import edgeR library
library(edgeR)
# we need to make our data into a DGE object 
dge <- DGEList(counts = data, group = tissues)
# normalize
dge.norm <- calcNormFactors(dge) 
# calculate filter and save it into a vector
isexpr <- rowSums(cpm(dge.norm) > cpm_threshold) >= cpm_needed
# use the vector to filter the genes
dge.norm.fltd <- dge.norm[isexpr, ]

```

It would be nice to get a quick visualization of our data, to see how our various tissues compare to each other. Of course, because we have so many genes, we can't easily look at them all at once, so we need to find a way to convert them into something simpler that captures most of their information. This involves a trick called dimensionality reduction. We're convertering our large-dimensional dataset (1 dimension for each gene), into a smaller number. One of the most common methods for doing this is Principle Component Analysis (PCA), which transforms our data into a number of principle components, which our composites of each of the genes, designed in such a way that they capture large fractions of the dataset's total variation. Then, we can look at the first and second principle components, and get a sense for how each of our samples are more or less similar to each other.

```{r pca}
# ggplot is the preimier plotting library in r
library(ggplot2)

# we'd like a table of normalizd counts per million to do our PCA on 
cpm = cpm(dge.norm.fltd)

# calculate principle components
pca_proc = prcomp(t(cpm), scale=T, center=T)
# make a dataframe with the pca results (pca_proc$x)
pca_res = data.frame(sample = rownames(pca_proc$x), pca_proc$x)
# make a column for organ for the dataframe
pca_res$organ = factor(sub('_.*_.*', '', pca_res$sample))

# plot
ggplot(pca_res, aes(PC2, PC1)) + 
    geom_point(aes(fill=organ), shape=21, color='black') +
    xlab(paste0('PC2: ', summary(pca_proc)$importance[2,2] * 100, '% variance')) +
    ylab(paste0('PC1: ', summary(pca_proc)$importance[2,1] * 100, '% variance'))

```

We use limma to do our actual linear modeling. Limma builds an initial model of the system using `lmFit()`. It uses design matrices to indicate how experimental conditions apply to each experimental group. One could imagine quite complex design matrices for quite complex experimental designs, but in a situation where you simply want to compare two groups, you can make something quite simple to indicate that each group has a single parameter associated with it.
We have 15 groups, not two, but in order to look for tissue specific genes, we need to look for genes which are differentially expressed between that tissue and all the other tissues. This means we'll be building 15 distinct models, one for each tissue.
We also use `voom()` to compute appropriate observational weights by transforming count data to log2-counts per million (log2-cpm) with associated weights, and estimating the mean-variance relationship, which prepares the data for linear modeling.
`topTable()` outputs a table of genes and associated modeling statistics.


```{r limma}
# a list in which we'll store our toptables
tops = list()

# loop through each tissue
for(tissue in unique(tissues)){
    # make our design matrix, using a vector saying if each sample is this tissue
    design <- model.matrix(~as.numeric(tissues == tissue))

    # voom to transform
    y <- voom(dge.norm.fltd, design, plot=F)
    # fit model for each gene
    fit <- lmFit(y, design)    
    
    # eBayes calculates necessary statistics
    fit = eBayes(fit)
    # get statistics for each gene
    tops[[tissue]] = topTable(fit, n=Inf)
    # it's useful to have the genes as a column as well as rownames
    tops[[tissue]]$gene = rownames(tops[[tissue]])
}

# dplyr and tidyr provide tools for manipulating dataframes
library(dplyr)
library(tidyr)
# we'd like one big dataframe of all the results
sig = bind_rows(tops, .id='organ') 
# we only want significantly upregulated genes, which have a pvalue < .01, and an lfc > 4
sig = filter(sig, adj.P.Val < .01, logFC > 4) 

# how many genes are there for each tissue?
summarize(group_by(sig, organ), n())

```

For each tissue, what proportion of their counts are from tissue-specific genes?

```{r tissue specific proportions}

# a melted dataframe is "tidy", and will be easier to work with
melted = melt(cpm(dge.norm))
colnames(melted) = c('gene', 'sample', 'counts')

# add an organ column
melted$organ = sub('_.*_.*', '', melted$sample)
# add a column specifying if the gene is tissue specific for the organ
melted$specific = paste0(melted$organ, melted$gene) %in% paste0(sig$organ, sig$gene)

# sum the number of counts that are specific or not specific for each organ
specific_counts = summarize(group_by(melted, organ, specific), counts = sum(counts))
# separate specific and not specific into two columns
specific_counts = spread(specific_counts, specific, counts)
# FALSE and TRUE are not great names to work with
specific_counts = rename(specific_counts, not_specific = `FALSE`, specific = `TRUE`)

# calculate total counts between specific and unspecific
specific_counts$total = specific_counts$not_specific + specific_counts$specific
# calculate percent of counts that are specific
specific_counts$specific_percent = specific_counts$specific / specific_counts$total * 100

# for visualization purposes we'll want to order our organs by how many counts are specific
# the forcats package makes ordering factors much easier
library(forcats)
# order organ
specific_counts$organ = fct_reorder(specific_counts$organ, specific_counts$specific_percent, .desc=T)

# plot!
ggplot(specific_counts, aes(organ, specific_percent)) +
    geom_col() + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust=.5)) + 
    ylab('Percent tissue specific') + 
    ggtitle('tissue specific')

```